WFH and broadband speed (title needs rework)

### 2020-12-21 paper/v1_sage/broadband.speed.covid.tex
new sage template. I need to change the path for the figures to lead to the new sage folder.

% interactcadsample.tex
% v1.03 - April 2017

\documentclass[]{interact}

\usepackage{epstopdf}% To incorporate .eps illustrations using PDFLaTeX, etc.
\usepackage{subfigure}% Support for small, `sub' figures and tables
%\usepackage[nolists,tablesfirst]{endfloat}% To `separate' figures and tables from text if required

\usepackage{natbib}% Citation support using natbib.sty
\bibpunct[, ]{(}{)}{;}{a}{}{,}% Citation support using natbib.sty
\renewcommand\bibfont{\fontsize{10}{12}\selectfont}% Bibliography support using natbib.sty

\theoremstyle{plain}% Theorem-like structures provided by amsthm.sty
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}

\theoremstyle{remark}
\newtheorem{remark}{Remark}
\newtheorem{notation}{Notation}

% see https://stackoverflow.com/a/47122900

\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\usepackage{dcolumn}
\def\tightlist{}

@@

\articletype{ARTICLE TEMPLATE}

@@

\author{\name{A. N. Author$^{a}$, John Smith$^{b}$}
\affil{$^{a}$Taylor \& Francis, 4 Park Square, Milton Park, Abingdon, UK; $^{b}$Institut für Informatik, Albert-Ludwigs-Universität, Freiburg, Germany}
}

\thanks{CONTACT A. N. Author. Email: \href{mailto:latex.helpdesk@tandf.co.uk}{\nolinkurl{latex.helpdesk@tandf.co.uk}}, John Smith. Email: \href{mailto:john.smith@uni-freiburg.de}{\nolinkurl{john.smith@uni-freiburg.de}}}

@@

\begin{keywords}
covid; internet; working from home; broadband speed; time-series
clusters
\end{keywords}

@@

The capability to work from home has previously been studied from the
perspective of whether work tasks in a given occupation both can be and
are allowed to be performed using digital technologies independently of
location or co-location with colleagues, including supervisors
\citep{allen2015effective, singh2013modeling}. However, successful
telecommuting also requires that the quality and reliability of digital
services, particularly home internet connection speeds, enable the
completion of work tasks with a minimum of delay or interruption. Prior
to the pandemic, the performance of broadband services with respect to
telecommuters was never tested at scale, as working from home and
connecting to colleagues and workplace resources via the internet was
the purview of a small minority of workers. Instead, leisure use in the
evening, when video streaming services are at their peak, has been used
to benchmark broadband performance and service delivery by different
Internet Service Providers (ISPs), at least in the UK \citep{ofcom2017}.
Yet the shift towards telecommuting during various stages of lockdown
around the world has been drastic and there are speculations that
post-Covid, the tendency to work from home will be much higher than
pre-Covid, raising questions around whether internet services can
accommodate the increased demand. For example, \(47\)\% of people in
employment in the UK worked soley from home in April \(2020\), whilst
the same figure only reached \(5\)\% the year before
\citep{ons2020, ons2020lm2019}. A back of the envelope calculation
suggests that up to 40\% of the working force could work from home on an
ongoing basis \citep{batty2020editorial}. Similar figures have been
reported for other countries \citep{felstead2020homeworking}. For
instance, \(37\)\% of the European workforce worked from home in April
\(2020\) with countries like Finland reaching \(60\)\%
\citep{eurofound2020}. In the US, almost half of the working population
worked from home during the same period because of the pandemic
\citep{brynjolfsson2020covid}, and a recent estimate indicated that
\(37\)\% of all jobs in the US can be permanently performed entirely
from home \citep{NBERw26948}.

None of these changes could have happened in the absence of reliable
information and communication technology (ICT) infrastructure -- both in
terms of software and hardware. But while software innovations are
easily diffused across space and society\footnote{See for example the
  huge success of videoconferencing apps such as Zoom
  \citep{marks2020zoom}.}, the same does not apply for ICT hardware
infrastructure such as internet broadband connectivity. The literature
describes first level digital divides in terms of the availability and
quality of internet connectivity, such as that manifest in different
geographies in the UK
\citep{riddlesden2014broadband, philip2017digital}. Second level digital
divides consider the presence or lack of the necessary skills to
effectively utilise digital technologies and the internet
\citep{blank2014dimensions, van2011internet}. The third level focuses on
the heterogenous returns of internet usage among different socioeconomic
groups and, consequently, how digital technologies can assist in
bridging or further enhancing existing socioeconomic divides.
\citep{stern2009levels, van2014digital, van2015third}. The capability to
telecommute is related to all three levels of digital divides, but more
importantly leads to differentiated outcomes regarding the economic
resilience of people and places to overcome a systemic shock such as the
current pandemic. We identify this as a new digital divide, one that
fundamentally alters the potential returns of internet use for the user
and wider community, assumes skills or functions that are present in
some occupations but not in others, and relies upon access to high
quality internet services. \textbf{WE NEED TO CONNECT OUR FINDINGS WITH
THIS AND JUSTIFY THAT THIS IS INDEED A NEW DIVIDE} As the quality of
internet infrastructure and services, as well as variation in
occupations are spatially dependent and clustered in space, our approach
offers a framework for understanding which types of places, are more
likely to land on the right side of this new digital divide. By asking
how resilient broadband speeds are as experienced in different parts of
the UK during a time of extreme demand, we interrogate which places
benefit from the greater economic resilience digital technologies can
offer, not only during the pandemic, but also into the future.

The structure of this paper is as follows. First we review the
literature on telecommuting and digital divides to better understand the
origins of the new digital divide revealed by the pandemic and its
impact on the economic resilience of different places. We then describe
our data and methodology. Our results section first offers a
classification of how internet services vary across the UK local
authorities and then assesses whether these clusters replicate or
repudiate other socio-economic and geographic patterns of economic
resilience.

@@

In this analysis, the terms `telecommuting' and `working from home' are
used interchangeably, as most remote labour during the Covid-19 crisis
was carried out in the homes of individual employees rather than any
other location \citep{eurofound2020}. However, it should be noted that
previous research has explored how telecommuting can occur in other
places, including satellite offices or on public transport
\citep{felstead2012rapid, siha2006telecommuting}. Previous research has
also used a variety of definitions to measure the level of telecommuting
within different workforces, distinguishing, for example, between those
directly employed, indirectly employed, self-employed, full-time or
part-time, and those who use digital technologies to work remotely
full-days or part-days
\citep{allen2015effective, bailey2002review, haddad2009examination}. No
matter the definition, the option and capability to telecommute or work
from home has never been equally distributed spatially or
socio-economically any more than different industries and employment
opportunities have. For example, studies from the United States, the
Netherlands, and the UK found that telecommuters are most likely to hold
professional, managerial, and technical occupations where the workforce
is better educated and wealthier, and that there is suppressed demand
among women and part-time workers
\citep{headicar2016move, peters2004employees, singh2013modeling}.
Opportunities for working from home during the current pandemic have
likewise not been equally spread across the workforce.
\citet{NBERw26948} indicated that in the US, managers, educators, as
well as those working in computer-related occupations, finance, and law
can easily work from home, and that occupations with opportunities to
telecommute are associated with higher earnings. This is not the case
for the workforce occupied in more spatially fixed occupations, from
farming, construction and manufacturing to hospitality and care
services. In the US, these occupations tend to be lower-income,
non-white, without a university degree, live in rental accommodation and
lack health insurance \citep{NBERw27085}. Similar trends can be observed
for other countries. For example, \(75\)\% of workers with tertiary
education worked from home in Europe during spring \(2020\), whilst only
\(34\)\% of workers with secondary education and \(14\)\% of those
primary education did so \citep{eurofound2020}.

@@

Our understanding of telecommuting as a product of enabled occupations
can be described as a manifestation of the third level digital divide,
as those who are able to use digital technologies to work from home
benefit from a high rate of return on their use of the internet in terms
of autonomy, flexibility, and time saved from commuting
\citep{peters2004employees, siha2006telecommuting, singh2013modeling}.
These returns have been even greater during the Covid-19 crisis, when
those with the ability to telecommute also have the ability to maintain
their employment whilst protecting their health. However, the success of
these arrangements has been dependent upon the first level digital
divide, which is associated with access and quality of internet
connectivity at a time of extremely high demand. \citet{SALEMINK2017360}
provides a systematic review of the pre-pandemic, first level digital
divide in infrastructure quality between urban and rural areas in
various advanced economies. Yet whether this variation in infrastructure
quality affects the spatial footprint of telecommuting has not
previously been measured. There are indications that those who purchase
high speed connections consume more data of all sorts and use their
connections for a variety of purposes \citep{hauge2011consumer}, and
that there is a correlation between access to internet services and a
reduction in household transport spend \citep{bris2017ict}. Whether the
implication is more internet use and less travel because of increased
telecommuting, these studies suggest that better internet services
enable households to make savings and efficiencies, an example of the
first level digital divide reinforcing the third level. Yet did the
purchase of high speed connections and increased internet access also
prepare households for long-term home-working, enforced by government
restrictions? The extreme demand during the pandemic provides a new
opportunity to understand how infrastructure accessibility, quality, and
reliability affects telecommuting, particularly in light of the high
volumes of bandwidth-intensive video conferencing required in order to
avoid the face-to-face contact that could increase the spread of
infection. We seek to answer how internet service resilience contributes
to or reduces economic resilience when the latter is dependent upon the
capability to work from home, and thus improve our understanding of this
new digital divide. \textbf{RELATED TO YOUR POINT ABOVE, WE MAY CONCLUDE
THIS IS NOT A NEW DIGITAL DIVIDE, BUT THAT IS STILL ANSWERING A USEFUL
QUESTION - MIGHT END UP REWORDING AS INTERACTION BETWEEN DIGITAL
DIVIDES? ET: OR REINFORCING THE 3RD LEVEL DIGITAL DIVIDE}

The multi-layered digital divides intersect with material divides and
the economic geography of the UK. Following the regional economic
resilience literature, which underlines the differentiated capacity of
cities and regions to escape or recover from economic crises
\citep{martin2012regional, kitsos2018economic}, different places have
different industrial and occupational profiles, and these affect the
aggregated potential capacity of places for telecommuting. Such profiles
are associated with longstanding inequalities in the UK and their
spatial representation as a North-South divide
\citep{martin_north_south}. Various studies have illustrated severe
inequalities between the north and the south part of the UK regarding,
for example, skills and human capital, unemployment, productivity and
prosperity \citep{lee2014grim, mccann2020perceptions, dorling2018peak}.
Some scholars have even argued that the UK suffers some of the highest
level of interregional inequalities in the global north
\citep{gal2018reducing, mccann2016uk}. Not only are all three levels of
digital divides associated to a certain extent with or shaped by the
geography of the UK, but the intersection of the digital and material
divides affects the capacity of places to overcome some of the economic
effects of the Covid-19 pandemic. Importantly, this is the first time
that digital technologies became an essential tool for economic
resilience for such a great part of the population.

@@

@@

Time-series clustering methods have been developed in order to deal with
clustering problems linked to, for instance, stock or other financial
data, economic, governmental or medical data as well as machine
monitoring
\citep{aggarwal2013time, aggarwal2001surprising, hyndman2015large, WARRENLIAO20051857}.
The main challenge -- and also the difference with cross-sectional
clustering problems -- is data dimensionality given the multiplicity of
data points for every individual object -- local authorities in our case
-- included in the data set. Time-series are dynamic data as the value
of the observations change as a function of time
\citep{aghabozorgi2015time}. This high dimensionality leads to (i)
computational and algorithmic challenges regarding handling these data
and building algorithms to perform clustering over long time-series, and
(ii) open questions regarding the choice of similarity measures in order
to cluster similar times series objects together considering the whole
length of the time-series and overcoming issues around noise, outliers
and shifts \citep{lin2004iterative, aghabozorgi2015time}.

Time-series clustering methods utilising the whole length of time-series
can be grouped in three categories. The first -- model-based approaches
-- is based on recovering the underlying model for each time-series and
then applying clustering algorithms on the model parameters of each
time-series \citep{aghabozorgi2015time}. The main criticism is the
cluster accuracy for nearby clusters \citep{mitsa2010temporal}. The
second approach is based the formation of vectors of features derived
from the original time-series. These new data of reduced dimensionality
are then clustered using conventional clustering algorithms.

For this paper we utilise the third category of time-series clustering
methods known as shape-based approaches. These methods match two sperate
time-series objects based on the similarity of their shapes through the
calculation of distances between the shapes, and are thus better
equipped to capture similarities between short length time-series
\citep{aghabozorgi2015time}. This approach serves best this paper
because (i) we identify clusters of UK local authorities with similar
temporal signatures -- i.e.~shapes -- of experienced internet speeds and
(ii) the length of our time-series is short (see the data discussion in
this section).

Another import element of time-series clustering is the actual
clustering algorithm. Similar to the clustering of cross-sectional data,
we can employ partitioning algorithms, which lead to non-overlapping
clusters, hierarchical clustering, which classifies clusters at
different levels, and fuzzy algorithms, which create overlapping
clusters \citep{sardatime}. Because of the simplicity of the
implementation and the interpretability of the results, we utilise here
partitioning clustering based on the widely used \emph{k-Means}
algorithm. This is an iterative algorithm, which begins with defining
the desired number of clusters \emph{k}. Then each observation is
randomly assigned to a cluster from the \([1,k]\) space. This initial
cluster assignment is followed by iterations in order to minimise the
distance between the centroids of the clusters and the observations
assigned to these clusters \citep{james2013introduction}.

There are a number of differences between the above described
application of \emph{k-Means} for cross-sectional data and its
application for times series data. Instead of creating clusters around
centroids, a common approach is to create clusters around
\emph{medoids}, which are representative time-series objects with a
minimal distance to all other cluster objects \citep{sardatime}. Also,
instead of calculating the Euclidean distance between centroids and data
points, more complex distance measures need to be employed in order to
capture the similarity between a time-series object and a medoid. A
common distance measure for shape-based time-series clustering is
Dynamic Time Warping (DTW). Using its underpinning dynamic programming
algorithm, DTW compares two time-series objects to find the optimum
warping path between them. DTW is widely used in order to overcome
limitations linked to the use of Euclidean distance
\citep{sardatime, berndt1994using, ratanamahatana2004everything}. The
\texttt{R} package \texttt{dtwclust} has been used for the time-series
clustering \citep{dtwclust}.

To assess the quality and reliability of internet across local
authorities in the UK during the time when the population were told to
work from home if at all possible we utilise unique data comprising
individual internet speed tests from Speedchecker Ltd\footnote{\url{https://www.broadbandspeedchecker.co.uk/}}.
This a private company that allows internet users to check their own
broadband upload and download speeds, and stores every speed-check with
timestamp and geolocation information. These data have been used before
to assess digital divides \citep{riddlesden2014broadband} and the impact
of local loop unbundling regulatory processes
\citep{nardotto2015unbundling}. By using volunteered geographic data, we
are able to assess the \emph{experienced} internet speed by users, which
may differ from the \emph{advertised} maximum speeds of Internet Service
Providers (ISPs). We are particularly interested in upload speeds and
the frequency of speed tests over the period from March to May 2020, as
government statements indicate this is when UK workers were told to work
from home if at all possible \textbf{{[}need ref?{]}}. Average upload
speeds are slower than average download speeds, at \(9.3\)Mb/s mean
upload speed for the whole sample during the period of interest,
compared to \(29.6\)Mb/s for download speeds, but they are also less
associated with internet-based, high-demand, leisure activities such as
video streaming. Therefore, upload speeds are more relevant to
work-related activities such as uploading documents or two-way audio,
video, and text-based communication systems. Meanwhile, the frequency of
speed tests was important in identifying the temporal profile which woud
give us most insight into experienced internet service and resilience,
and provide an indication of the volume of experience over particular
units of time.

The first step in the workflow after dropping some outliers following
\citet{riddlesden2014broadband} was to transform the individual,
geolocated and time-stamped tests to more meaningful aggregates both in
terms of space and time. The frequency of testing indicates that whilst
there is an overall trend of increased testing \textbf{or increased
experience of internet service NOT SURE WHAT THIS MEAND} from March to
April and then a slight reduction from April to May, this trend masks
substantial variation by not only the day of the week, but also time of
day, as can be seen in Figure \ref{test2020}. Thus, a daily aggregation
of upload speeds would mask the variation in experience over the course
of each weekday. Furthermore, the importance of this variation is
highlighted by a comparison with the same period in \(2019\), as in
Figure \ref{test2019}, when the volume of testing and thus of experience
of internet service quality was much more tied to the evening leisure
peak and presumably to download speeds. Since the increase in testing
during the working day in \(2020\) is an indication that users have a
greater perception of the variable quality of internet service,
particularly during a new morning peak of testing for service
reliability, we decided to include a measure of hourly variation in our
temporal profiles. However, although this is a large data set
(\(241,088\) individual tests during the working hours of study period),
there are not enough observations for each Local Authority District
(LAD) and for each working hour of each working day (\(631\) speed tests
per LAD on average) to profile speeds at that level of detail.
Therefore, we aggregate all the speed-checks during the \(13\) weeks of
March to May inclusive for weekdays in \(2020\) by each hour of the day
and day of the week. As our research aims to identify the geography of
internet service resilience for work purposes, bank holidays and the
hours between midnight and 6:00 were excluded, as well as weekend days.
The composite week time-series thus comprise \(18\) hours multiplied by
\(5\) weekdays or \(90\) time points per series. We also aggregate these
data spatially because we could not follow individuals or households and
connect data points. The time-series were calculated for each of the
\(382\) LADs in the UK, standardised, and then a \emph{k}-means
partitioning around medoids clustering algorithm was applied using DTW.
We initially run the algorithm for \(k \in \mathbb{N} \bigcap [5,15]\)
and used cluster validity indices (CVIs) to pick the optimal solution of
\(k = 13\). Following \citet{sardatime} the majority vote for the
following CVIs was used: Silhouette (max), Score function (max),
Calinski-Harabasz (max), Davies-Bouldin (min), Modified Davies-Bouldin
(DB*, min), Dunn (max), COP (min).

@@

We then review the characteristics of each cluster, including number of
LADs, and the temporal profile of upload speed by hour of the day and
day of the composite week. Since the quality and reliability of internet
services vary in time and space due to both supply and demand-side
influences, we use a number of different measures to then describe
experienced upload speeds per cluster. These include: a) mean,
experienced connection speed, b) standard deviation or the amount of
fluctuation from the mean, and c) the variation in speeds at particular
times of day when working from home is more likely to take place. We
take account of all three measurements in our descriptives statistics of
upload speeds in order to determine how resilient broadband speeds are
as experienced in different parts of the UK during a time of extreme
demand.

\textbf{More data details and some figures, descriptive stats}

>>>>

\documentclass[Royal,times,sageh]{sagej}

\usepackage{moreverb,url,natbib, multirow, tabularx}
\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}

\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\usepackage{dcolumn}
\def\tightlist{}
\newcommand{\beginappendix}{\setcounter{table}{0} \renewcommand{\thetable}{A\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{A\arabic{figure}}}

@@

@@

\runninghead{}

\author{A. Uthor*\affilnum{1,2}, O. Tro\affilnum{2}, O. Vriga\affilnum{3}}

\affiliation{\affilnum{1}{Department of Incredible Research, University A, City A, Country A}\\\affilnum{2}{Department of Applied Things, University B, City B, Country B}\\\affilnum{3}{Very Important Stuff Committee, Institute C, City C, Country C}}

\corrauth{Corresponding author name, This is sample corresponding address.}

\email{\href{mailto:correspondingauthor@protonmail.com}{\nolinkurl{correspondingauthor@protonmail.com}}}

\keywords{covid; internet; working from home; broadband speed; time-series
clusters;}

@@

@@

The capability to work from home has previously been studied from the
perspective of whether work tasks in a given occupation both can be and
are allowed to be performed using digital technologies independently of
location or co-location with colleagues, including supervisors
\citep{allen2015effective, singh2013modeling}. However, successful
telecommuting also requires that the quality and reliability of digital
services, particularly home internet connection speeds, enable the
completion of work tasks with a minimum of delay or interruption. Prior
to the pandemic, the performance of broadband services with respect to
telecommuters was never tested at scale, as working from home and
connecting to colleagues and workplace resources via the internet was
the purview of a small minority of workers. Instead, leisure use in the
evening, when video streaming services are at their peak, has been used
to benchmark broadband performance and service delivery by different
Internet Service Providers (ISPs), at least in the UK \citep{ofcom2017}.

Yet the shift towards telecommuting during various stages of lockdown
around the world has been drastic and there are speculations that
post-Covid, the tendency to work from home will be much higher than
pre-Covid, raising questions around whether internet services can
accommodate the increased demand. For example, \(47\)\% of people in
employment in the UK worked solely from home in April \(2020\), whilst
the same figure only reached \(5\)\% the year before
\citep{ons2020, ons2020lm2019}. A back of the envelope calculation
suggests that up to 40\% of the working force could work from home on an
ongoing basis \citep{batty2020editorial}. Similar figures have been
reported for other countries \citep{felstead2020homeworking}. For
instance, \(37\)\% of the European workforce worked from home in April
\(2020\) with countries like Finland reaching \(60\)\%
\citep{eurofound2020}. In the US, almost half of the working population
worked from home during the same period because of the pandemic
\citep{brynjolfsson2020covid}, and a recent estimate indicated that
\(37\)\% of all jobs in the US can be permanently performed entirely
from home \citep{NBERw26948}.

None of these changes could have happened in the absence of reliable
information and communication technology (ICT) infrastructure -- both in
terms of software and hardware. But while software innovations are
easily diffused across space and society\footnote{See for example the
  huge success of videoconferencing apps such as Zoom
  \citep{marks2020zoom}.}, the same does not apply for ICT hardware
infrastructure such as internet broadband connectivity. The literature
describes first level digital divides in terms of the availability and
quality of internet connectivity, such as that manifest in different
geographies in the UK
\citep{riddlesden2014broadband, philip2017digital}. Second level digital
divides consider the presence or lack of the necessary skills to
effectively utilise digital technologies and the internet
\citep{blank2014dimensions, van2011internet}. The third level focuses on
the heterogenous returns of internet usage among different socioeconomic
groups and, consequently, how digital technologies can assist in
bridging or further enhancing existing socioeconomic divides.
\citep{stern2009levels, van2014digital, van2015third}. The capability to
telecommute is related to all three levels of digital divides, but more
importantly leads to differentiated outcomes regarding the economic
resilience of people and places to overcome a systemic shock such as the
current pandemic. The extreme level of demand for telecommuting
fundamentally alters the potential returns of internet use for the user
and wider community, assumes skills or functions that are present in
some occupations but not in others, and relies upon access to high
quality internet services.

As the quality of internet infrastructure and services, as well as
variation in occupations are spatially dependent and clustered in space,
our approach offers a framework for understanding the impact of and
interactions between the different levels of digital division in
different places with different characteristics. By asking how resilient
broadband speeds, and particularly upload speeds are as experienced in
different parts of the UK during a time of extreme demand, we
interrogate which places benefit from the greater economic resilience
digital technologies can offer, not only during the pandemic, but also
into the future. The structure of this paper is as follows. First we
review the literature on telecommuting and digital divides to better
understand the structural and spatial development of these practices
pre-pandemic, and thus their importance to the economic resilience of
different places. We then describe our data and methodology. Our results
section first offers a classification of how internet services vary
across the UK local authorities and then assesses whether these clusters
replicate or repudiate other socio-economic and geographic patterns of
economic resilience. The paper ends with a conclusions section.

@@

In this analysis, the terms `telecommuting' and `working from home' are
used interchangeably, as most remote labour during the Covid-19 crisis
was carried out in the homes of individual employees rather than any
other location \citep{eurofound2020}. However, it should be noted that
previous research has explored how telecommuting can occur in other
places, including satellite offices or on public transport
\citep{felstead2012rapid, siha2006telecommuting}. Previous research has
also used a variety of definitions to measure the level of telecommuting
within different workforces, distinguishing, for example, between those
directly employed, indirectly employed, self-employed, full-time or
part-time, and those who use digital technologies to work remotely
full-days or part-days
\citep{allen2015effective, bailey2002review, haddad2009examination}. No
matter the definition, the option and capability to telecommute or work
from home has never been equally distributed spatially or
socio-economically any more than different industries and employment
opportunities have. For example, studies from the United States, the
Netherlands, and the UK found that telecommuters are most likely to hold
professional, managerial, and technical occupations where the workforce
is better educated and wealthier, and that there is suppressed demand
among women and part-time workers
\citep{headicar2016move, peters2004employees, singh2013modeling}.

Opportunities for working from home during the current pandemic have
likewise not been equally spread across the workforce.
\citet{NBERw26948} indicated that in the US, managers, educators, as
well as those working in computer-related occupations, finance, and law
can easily work from home, and that occupations with opportunities to
telecommute are associated with higher earnings. This is not the case
for the workforce occupied in more spatially fixed occupations, from
farming, construction and manufacturing to hospitality and care
services. In the US, these occupations tend to be lower-income,
non-white, without a university degree, live in rental accommodation and
lack health insurance \citep{NBERw27085}. Similar trends can be observed
for other countries. For example, \(75\)\% of workers with tertiary
education worked from home in Europe during spring \(2020\), whilst only
\(34\)\% of workers with secondary education and \(14\)\% of those
primary education did so \citep{eurofound2020}.

@@

Our understanding of telecommuting as a product of enabled occupations
can be described as a manifestation of the third level digital divide,
as those who are able to use digital technologies to work from home
benefit from a high rate of return on their use of the internet in terms
of autonomy, flexibility, and time saved from commuting
\citep{peters2004employees, siha2006telecommuting, singh2013modeling}.
These returns have been even greater during the Covid-19 crisis, when
those with the ability to telecommute also have the ability to maintain
their employment whilst protecting their health. However, the success of
these arrangements has been dependent upon the first level digital
divide, which is associated with access and quality of internet
connectivity at a time of extremely high demand. \citet{SALEMINK2017360}
provides a systematic review of the pre-pandemic, first level digital
divide in infrastructure quality between urban and rural areas in
various advanced economies. Rural areas, predictably, fare worse. Yet
whether this variation in infrastructure quality affects the spatial
footprint of telecommuting has not previously been measured, in part
because telecommuting has not previously been the cause of greatest
demand and pressure on internet services.

There are indications that those who purchase high speed connections
consume more data of all sorts and use their connections for a variety
of purposes \citep{hauge2011consumer}, and that there is a correlation
between access to internet services and a reduction in household
transport spend \citep{bris2017ict}. Whether the implication is more
internet use and less travel because of increased telecommuting, these
studies suggest that better internet services enable households to make
savings and efficiencies, an example of the first level digital divide
reinforcing the third level. Yet did the purchase of high speed
connections and increased internet access also prepare households for
long-term home-working, enforced by government restrictions? The extreme
demand during the pandemic provides a new opportunity to understand how
infrastructure accessibility, quality, and reliability affects
telecommuting, particularly in light of the high volumes of
bandwidth-intensive video conferencing required in order to avoid the
face-to-face contact that could increase the spread of infection. We
seek to answer how internet service resilience contributes to or reduces
economic resilience when the latter is dependent upon the capability to
work from home. We also aim to improve our understanding of the impact
of first level digital division on telecommuting, and whether this
results in much more fundamental third level digital division than has
previously been perceived.

Furthermore, these multi-layered digital divides intersect with material
divides and the economic geography of the UK. Following the regional
economic resilience literature, which underlines the differentiated
capacity of cities and regions to escape or recover from economic crises
\citep{martin2012regional, kitsos2018economic}, different places have
different industrial and occupational profiles, and these affect the
aggregated potential capacity of places for telecommuting. Such profiles
are associated with longstanding inequalities in the UK and their
spatial representation as a North-South divide
\citep{martin_north_south}. Various studies have illustrated severe
inequalities between the north and the south regions of England
\textbf{replace `UK'} regarding, for example, skills and human capital,
unemployment, productivity and prosperity
\citep{lee2014grim, mccann2020perceptions, dorling2018peak}. Some
scholars have even argued that the UK suffers some of the highest level
of interregional inequalities in the global north
\citep{gal2018reducing, mccann2016uk}. Not only are all three levels of
digital divides associated, to a certain extent, with or shaped by the
geography of the UK, but the intersection of the digital and material
divides affects the capacity of places to overcome some of the economic
effects of the Covid-19 pandemic. Importantly, this is the first time
that digital technologies became an essential tool for economic
resilience for such a great part of the population.

@@

\hypertarget{sec:3.1}{%
\subsection{Time-Series Clustering}\label{sec:3.1}}

@@

Time-series clustering methods have been developed in order to deal with
clustering problems linked to, for instance, stock or other financial
data, economic, governmental or medical data as well as machine
monitoring
\citep{aggarwal2013time, aggarwal2001surprising, hyndman2015large, WARRENLIAO20051857}.
The main challenge -- and also the difference with cross-sectional
clustering problems -- is data dimensionality given the multiplicity of
data points for every individual object, local authorities in our case,
included in the data set. Time-series are dynamic data as the value of
the observations change as a function of time
\citep{aghabozorgi2015time}. This high dimensionality leads to (i)
computational and algorithmic challenges regarding handling these data
and building algorithms to perform clustering over long time-series, and
(ii) open questions regarding the choice of similarity measures in order
to cluster similar times series objects together considering the whole
length of the time-series and overcoming issues around noise, outliers
and shifts \citep{lin2004iterative, aghabozorgi2015time}.

For this paper we utilise a category of time-series clustering methods
known as shape-based approaches. These methods match two sperate
time-series objects based on the similarity of their shapes through the
calculation of distances between the shapes, and are thus better
equipped to capture similarities between short length time-series
\citep{aghabozorgi2015time}. This approach serves best this paper
because (i) we identify clusters of UK local authorities with similar
temporal signatures -- i.e.~shapes -- of experienced internet speeds and
(ii) the length of our time-series is short (see the data discussion in
this section).

Another import element of time-series clustering is the actual
clustering algorithm. Similar to the clustering of cross-sectional data,
we can employ partitioning algorithms, which lead to non-overlapping
clusters, hierarchical clustering, which classifies clusters at
different levels, and fuzzy algorithms, which create overlapping
clusters \citep{sardatime}. Because of the simplicity of the
implementation and the interpretability of the results, we utilise here
partitioning clustering based on the widely used \emph{k}-means
algorithm. This is an iterative algorithm, which begins with defining
the desired number of clusters \emph{k}. Then each observation is
randomly assigned to a cluster from the \([1,k]\) space. This initial
cluster assignment is followed by iterations in order to minimise the
distance between the centroids of the clusters and the observations
assigned to these clusters \citep{james2013introduction}.

There are a number of differences between the above described
application of \emph{k}-means for cross-sectional data and its
application for times series data. Instead of creating clusters around
centroids, a common approach is to create clusters around
\emph{medoids}, which are representative time-series objects with a
minimal distance to all other cluster objects \citep{sardatime}. Also,
instead of calculating the Euclidean distance between centroids and data
points, more complex distance measures need to be employed in order to
capture the similarity between a time-series object and a medoid. A
common distance measure for shape-based time-series clustering is
Dynamic Time Warping (DTW). Using its underpinning dynamic programming
algorithm, DTW compares two time-series objects to find the optimum
warping path between them. DTW is widely used in order to overcome
limitations linked to the use of Euclidean distance
\citep{sardatime, berndt1994using, ratanamahatana2004everything}. The
\texttt{R} package \texttt{dtwclust} has been used for the time-series
clustering \citep{dtwclust}.

\hypertarget{sec:3.2}{%
\subsection{Experienced Broadband Speeds}\label{sec:3.2}}

To assess the quality and reliability of internet across local
authorities in the UK during the time when the population were told to
work from home if at all possible we utilise unique data comprising
individual internet speed tests from Speedchecker Ltd\footnote{\url{https://www.broadbandspeedchecker.co.uk/}}.
This is a private company that allows internet users to check their own
broadband upload and download speeds, and stores every speed-check with
timestamp and geolocation information. These data have been used before
to assess digital divides \citep{riddlesden2014broadband} and the impact
of local loop unbundling regulatory processes
\citep{nardotto2015unbundling}. By using volunteered geographic data, we
are able to assess the \emph{experienced} internet speed by users, which
may differ from the \emph{advertised} maximum speeds of ISPs.

We are particularly interested in upload speeds and the frequency of
speed tests over the period from March to May \(2020\), as government
statements indicate this is when UK workers were told to work from home
if at all possible \citep{GovUK2020}. Average upload speeds are slower
than average download speeds, at \(9.3\)Mb/s mean upload speed for the
whole sample during the period of interest, compared to \(29.6\)Mb/s for
download speeds, but they are also less associated with internet-based,
high-demand, leisure activities such as video streaming. Therefore,
upload speeds are more relevant to work-related activities such as
uploading documents or two-way audio, video, and text-based
communication systems. Meanwhile, the frequency of speed tests was
important in identifying the temporal profile which would give us most
insight into experienced internet service and resilience, and provide an
indication of the volume of experience over particular units of time.

The first step in the workflow after dropping some outliers following
\citet{riddlesden2014broadband} was to transform the individual,
geolocated and time-stamped tests to more meaningful aggregates both in
terms of space and time. The frequency of testing indicates that whilst
there is an overall trend of increased testing from March to April and
then a slight reduction from April to May, this trend masks substantial
variation by not only the day of the week, but also time of day, as can
be seen in Figure \ref{test2020}. Thus, a daily aggregation of upload
speeds would mask the variation in experienced service over the course
of each weekday. Furthermore, the importance of this variation is
highlighted by a comparison with the same period in \(2019\), as in
Figure \ref{test2019}, when the volume of testing and thus of experience
of internet service quality was much more tied to the evening leisure
peak and presumably to download speeds. Since the increase in testing
during the working day in \(2020\) is an indication that users have a
greater perception of the variable quality of internet service,
particularly during a new morning peak of testing for service
reliability, we decided to include a measure of hourly variation in our
temporal profiles.

However, although this is a large data set -- \(241,088\) individual
tests performed during weekday hours across the study period -- there
are not enough observations for each Local Authority District (LAD) and
for each working hour of each working day -- \(631\) speed tests per LAD
on average -- to profile speeds at that level of detail. Therefore, we
aggregate all the speed-checks during the \(13\) weeks of March to May
inclusive for weekdays in \(2020\) by each hour of the day and day of
the week. As our research aims to identify the geography of internet
service resilience for work purposes, bank holidays and the hours
between midnight and \(6:00\) were excluded, as well as weekend days.
The composite week time-series thus comprise \(18\) hours multiplied by
\(5\) weekdays or \(90\) time points per series. We also aggregate these
data spatially because we could not follow individuals or households and
connect data points. The time-series were calculated for each of the
\(382\) LADs in the UK, standardised, and then a \emph{k}-means
partitioning around medoids clustering algorithm was applied using DTW.
We initially run the algorithm for \(k \in \mathbb{N} \bigcap [5,15]\)
and used cluster validity indices (CVIs) to pick the optimal solution of
\(k = 13\). Following \citet{sardatime} the majority vote for the
following CVIs was used: Silhouette (max), Score function (max),
Calinski-Harabasz (max), Davies-Bouldin (min), Modified Davies-Bouldin
(DB*, min), Dunn (max), COP (min).

@@

In Section \protect\hyperlink{sec:4.1}{4.1}, we review the temporal
profile of upload speed by hour of the day and day of the composite
week, as well as the experienced speed characteristics of each cluster.
Since the quality and reliability of internet services vary in time and
space due to both supply and demand-side influences, we use a number of
different measures to describe experienced upload speeds per cluster.
These include: a) mean, experienced connection speed, b) standard
deviation or the amount of fluctuation from the mean, and c) the
variation in speeds at particular times of day when working from home is
more likely to take place. We take account of all three measurements in
our descriptive statistics of upload speeds in order to determine how
resilient broadband speeds are as experienced in different parts of the
UK during a time of extreme demand.
